---
title: "3 STEPS pipeline  for Identification & Visualization of Significant Methylation Susceptible Loci (MSL)"
author: "Lakshay Anand"
date: "2/15/2019"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


<hr>
```{r klippy, echo=FALSE, include=TRUE}
klippy::klippy(color = '#328913',tooltip_message = 'Click here to copy!', tooltip_success = 'Copied Succesfully!')
```


<style>
a { color:#328913;}
a:hover{color:black; 
text-decoration:none;}


</style>

##CONTENTS
[What the pipeline does?](#what)

[What you need?](#need)

[Getting Started](#getstart)

[Function Sourcecode](#funso)




<hr>

```{r echo=FALSE}

####################################################################
########### STATISTICAL ANALYSIS FOR FINDING SIGNIFICANT LOCI BW GROUPS####
############## @author: Lakshay Anand ###############


find.sig.loci<- function(msap.input){
  
  out=list()
  
  ptrn=msap.input[[2]]
  grpNAME=c()
  grpCOUNT=c()
 
  grpALL=data.frame()
  for(d in 1:length(ptrn)){
    
    Grp=ptrn[[d]]
    
    rownames(Grp)= c(paste("G",d,"-",c(1:nrow(Grp)),sep = ""))
    grpNAME[d]=paste("G",d,sep = "")
    grpCOUNT[d]=nrow(Grp)
    grpALL=rbind.data.frame(grpALL,Grp)
    
    
  }
  
  
  cat("analysing.... \n")
  cat("Total loci:",ncol(grpALL),"\n")
  cat("Total samples:",nrow(grpALL),"\n")
  cat("Total number of groups:",length(ptrn),"\n")
  
  
  ############## replace types with numbers
  message("replacing types with numbers suitable for fisher test and heatmap.. \n")
  for(i in 1:ncol(grpALL)){
    
    
    for(j in 1:nrow(grpALL)){
      
      if(grpALL[j,i]=="u"){
        
        grpALL[j,i]=1
      }
      else if(grpALL[j,i]=="h"){
        
        grpALL[j,i]=2
      } else if(grpALL[j,i]=="i"){
        
        grpALL[j,i]=3
      } else if(grpALL[j,i]=="f"){
        
        grpALL[j,i]=4
      }
      
      
      
    }
    
    
    
  }
  
  for(i in 1:ncol(grpALL)){
    
    grpALL[,i]=as.integer(grpALL[,i])
    
  }
  
  
  
  #####################################################
  #creating groups vector for fisher test
  groups.vector=c()
  
  for(i in 1:length(rownames(grpALL))){
    
    groups.vector[i]=strsplit(rownames(grpALL)[i],"-")[[1]][1]
  }
  
  
  ############### multiple fisher test ##########
  message("Performing multiple fisher tests \n")
  pval=c()
  loci=c()
  for(i in 1:ncol(grpALL)){
    
    loci[i]=colnames(grpALL)[i]
    
    pval[i]=tryCatchFisher(table(groups.vector,grpALL[,i]))
    
    
  }
  
  
  message("Calculating FDR using benjamini and hochberg \n")
  #result of the fisher test wit p-vale adjustment using benjamini
  
  grpALL_LOCI=data.frame(loci,pval,p.adjust(pval,method = "hochberg"))
  
  message("filtering significant loci \n")
  ### filter the loci with FDR <0.05
  grp_sigloci=grpALL_LOCI[grpALL_LOCI[,3]<0.05,]
  
  
  
  ###collect the significant loci
  sig.loci=as.character(grp_sigloci[,1])
  sig.loci=na.omit(sig.loci)
  cat(length(sig.loci)," loci are found significant. \n")
  ##### filter the loci from the matrix
  
  grpALL_with_sig_loci=grpALL[,sig.loci]
  
  
  out[[1]]=as.matrix(grpALL_with_sig_loci)
  out[[2]]=sig.loci
  out[[3]]=grpALL_LOCI
  out[[4]]=grpALL
  out[[5]]=data.frame(grpNAME,grpCOUNT)
  names(out)=c("heatmap.matrix","Significant.Loci","All.loci.results","complete.data","group.cont")
  return(out)
  }




#### the fisher test in try catch
tryCatchFisher<-function(v){
  tryCatch(fisher.test(v)[[1]],
           error = function(e) {"na"})
}




```

>## What the pipeline does?<a name="what"></a>

The pipeline will help you  find and visualize significant Methylation Susceptible Loci (MSL) between the groups using 3 SIMPLE STEPS:

* Find the MSL using `msap` package.
* Find the significant loci using a custom R function `find.sig.loci( )` that will take the output of `msap` and return the significant loci. The function is based on multiple fisher exact tests followed by p-value adjustment using FDR.
* Plot significant MSL as heatmap depicting the four types of loci viz.Type I-IV.Type I, unmethylated; Type II, inner cytosine methylation; Type III, hemi-methylation of the outer cytosine; Type IV, fully methylation. <a name="need"></a>



>## What do you need?

You will need to install the package `msap` and prepare the **data format** that it needs. That is all you need for this pipeline. <a name="getstart"></a>

>## Getting Started

Lets run the three steps for the pipeline. Before starting with the following command, copy and run the function sourcecode provided in the last section.

###STEP 1

Use the `msap` package and the input data to find the MSL.

```{r eval=FALSE}
library(msap)
msap.out=msap("your_data.csv",name = "any_name")
```

```{r echo=FALSE,results='hide'}
library(msap)
msap.out=msap("BO_GNTYPE.csv",name = "OG")
```

###STEP 2

pass the output to the custom function `find.sig.loci( )` (more info about the function in subsequent sections)

```{r eval=TRUE }

my.out=find.sig.loci(msap.out)
```


Note: The warning during coercion to NA is produced when the input data has only one variable and can't be used for fisher test. The cistome function returns a na for such cases.


and now the output `my.out` is a list of five elements:

```{r eval=TRUE }

names(my.out)
```

where,
*`heatmap.matrix` is a output matrix of numbers significant loci suitable for creating the heatmap.

*`significant.loci` is character vector containing the significant loci selected.

*`All.loci.results` is the overall result showing the p-values and adjusted p-values for all the loci(not filtered). The pipeline uses a filter of FDR<0.05. While using this data you can use your own filter.

*`complete.data` contains the matrix of all loci and samples together.

*`group.count` can be used to know the sample count for each group for using it in heatmap cluster coloring.


```{r }
#significant loci
my.out$Significant.Loci
```

```{r}
head(my.out$All.loci.results)
```

```{r }
#number of samples
my.out$group.cont
```


###STEP3

Now get the heatmap matrix data and plot the heatmap. I have used the `heatmap.2( )` function from `gplots` package. You can use any package or settings.

```{r }
library(gplots)
heat.data=my.out$heatmap.matrix

#for getting the group colors for the samples marked as G1,G2.. size of which obtained from group,cont dataframe
row.cluster.colors= c(rep("green",21),rep("red",20))
heatmap.2(as.matrix(heat.data),
                        Rowv = F,
                        trace = "none",
                        col = c('#eff3ff','#bdd7e7','#6baed6','#2171b5') ,
                        RowSideColors = ,
                        dendrogram="column")
```


NOTE: The four shades of blue corresponds to the four levels of loci TypeI-IV(Type I, unmethylated; Type II, inner cytosine methylation; Type III, hemi-methylation of the outer cytosine; Type IV, fully methylation.) in the order 1-4. 





and its done!<a name="funso"></a>

>## Function Sourcecode

Copy the following function code and run it in your R environment. This will create two functions, `find.sig.loci( )` and `tryCatchFisher( )`. 

```{r eval=FALSE}

###########################################################################
########### STATISTICAL ANALYSIS FOR FINDING SIGNIFICANT LOCI BW GROUPS####
############## @author: Lakshay Anand #####################################


find.sig.loci<- function(msap.input){
  
  out=list()
  
  ptrn=msap.input[[2]]
  grpNAME=c()
  grpCOUNT=c()
  grpALL=data.frame()
  for(d in 1:length(ptrn)){
    
    Grp=ptrn[[d]]
    rownames(Grp)= c(paste("G",d,"-",c(1:nrow(Grp)),sep = ""))
    grpNAME[d]=paste("G",d,sep = "")
    grpCOUNT[d]=nrow(Grp)
    grpALL=rbind.data.frame(grpALL,Grp)
    
  }
  
  
  cat("analysing.... \n")
  cat("Total loci:",ncol(grpALL),"\n")
  cat("Total samples:",nrow(grpALL),"\n")
  cat("Total number of groups:",length(ptrn),"\n")
  
  
  ############## replace types with numbers
  message("replacing types with numbers suitable for fisher test and heatmap.. \n")
  for(i in 1:ncol(grpALL)){
    
    
    for(j in 1:nrow(grpALL)){
      
      if(grpALL[j,i]=="u"){
        
        grpALL[j,i]=1
      }
      else if(grpALL[j,i]=="h"){
        
        grpALL[j,i]=2
      } else if(grpALL[j,i]=="i"){
        
        grpALL[j,i]=3
      } else if(grpALL[j,i]=="f"){
        
        grpALL[j,i]=4
      }
      
      
      
    }
    
    
    
  }
  
  for(i in 1:ncol(grpALL)){
    
    grpALL[,i]=as.integer(grpALL[,i])
    
  }
  
  
  
  #####################################################
  #creating groups vector for fisher test
  groups.vector=c()
  
  for(i in 1:length(rownames(grpALL))){
    
    groups.vector[i]=strsplit(rownames(grpALL)[i],"-")[[1]][1]
  }
  
  
  ############### multiple fisher test ##########
  message("Performing multiple fisher tests \n")
  pval=c()
  loci=c()
  for(i in 1:ncol(grpALL)){
    
    loci[i]=colnames(grpALL)[i]
    
    pval[i]=tryCatchFisher(table(groups.vector,grpALL[,i]))
    
    
  }
  
  
  message("Calculating FDR using benjamini and hochberg \n")
  #result of the fisher test wit p-vale adjustment using benjamini
  
  grpALL_LOCI=data.frame(loci,pval,p.adjust(pval,method = "hochberg"))
  
  message("filtering significant loci \n")
  ### filter the loci with FDR <0.05
  grp_sigloci=grpALL_LOCI[grpALL_LOCI[,3]<0.05,]
  
  
  
  ###collect the significant loci
  sig.loci=as.character(grp_sigloci[,1])
  sig.loci=na.omit(sig.loci)
  cat(length(sig.loci)," loci are found significant. \n")
  ##### filter the loci from the matrix
  
  grpALL_with_sig_loci=grpALL[,sig.loci]
  
  
  out[[1]]=as.matrix(grpALL_with_sig_loci)
  out[[2]]=sig.loci
  out[[3]]=grpALL_LOCI
  out[[4]]=grpALL
  out[[5]]=data.frame(grpNAME,grpCOUNT)
  names(out)=c("heatmap.matrix","Significant.Loci","All.loci.results","complete.data","group.cont")
  
  return(out)
  }




#### the fisher test in try catch
tryCatchFisher<-function(v){
  tryCatch(fisher.test(v)[[1]],
           error = function(e) {"na"})
}




```






